Toward theories of deliberated preferences

### 2020-10-20 simple.tex
Meilleure intro

\section{Motivation} 
\commentYM{dans la mesure où tu dis concevoir ou avoir conçu ce draft comme étant presque soumettable, pourquoi as-tu cette section "motivation" plutôt qu'une classique introduction ?}
It has escaped nobody’s attention that 
studying reflective preferences, in supplement to shallow preferences, is an important endeavor for economics: we want to recommend to individual decision makers; we want to adopt optimization criteria for society, and this ought not to merely faithfully reproduce unreflectful behavior.\commentYM{je trouve cette entame provocatrice assez maladroite, car elle s'appuie sur l'opposition reflective vs. shallow preference, qui ne parle (je pense) pas immédiatement au lecteur. Je pense indispensable de définir cette dichotomie avant d'en parler. Par ailleurs que veut dire "is an important endeavor for economics" ? Il me semble au contraire que c'est une tâche qui devrait tenir à coeur aux économistes, mais qui pourtant n'est pas pour eux une tâche à laquelle ils s'attèlent vigoureusements. Enfin, l'anglais de cette phrase me semble un peu maladroit} This is one important difference between the normative and the (inappropriately called, as will be seen) “descriptive” approach in economics\commentYM{où discutes-tu cette distinction, dans son application à l'économie en particulier ?}. As \citet{hausman_preference_1994} illustrate with an example of a person drinking a fatal poison while erroneously thinking that it is water, shallow preferences are not sufficient to inform normative economics.
Yet, classical methods to obtain preference models in economics are either formally defined on the basis of observable data that do not describe reflective preferences, or are obtained by other means than observable preferential data and systematic reproducible formal procedures. \commentYM{je devine ce que tu veux dire, mais je ne comprends pas la dernière phrase. Il me semble important que tu t'expliques ici, car je pense que le lecteur ici ne peut pas comprendre la niche dans laquelle s'inscrit ton projet}

This article is interested in defining theories of decision making based on another kind of observable data\commentYM{other than what? Juste avant tu dis que les autres s'appuient sur des données qui ne reflètent pas les préférences qui t'intéressent, et maintenant tu dis que tu vas regarder autre chose qu'eux. Cela donne l'impression que tu tournes autour du pot, car tu ne dis pas ce que les autres regardes et pourquoi ça ne t'intéresse pas}, namely, the reactions of decision makers to arguments. This permits to overcome a current important limitation of normative approaches, that of being not empirical (in a sense that will be developed below).\commentYM{phrase incompréhensible pour le lecteur, puisque tu dis toi même que tu en expliqueras les termes plus tard. C'est une rhétorique difficile à tenir dans une introduction}

By reflective preferences (of an individual), I mean here preferences that represent informed, pondered decisions, that would be taken after due consideration by the individual. Reflective preferences can be contrasted to shallow preferences, those that \citet[p.\ 16]{von_neumann_theory_2004} describe as “intuitive”, as originating from an “immediate sensation” (cited in \citet{fishburn_retrospective_1989}).
This distinction draws on our capacity, as human beings, to think about our shallow preferences and consider some of them as mistaken \citep{frankfurt_freedom_1971}. \commentYM{ces définitions devraient être mises + haut, car le lecteur ne peut pas comprendre ce que tu racontes sans elles}A famous example of such a situation is the one of \citet[pp.\ 101--103]{savage_foundations_1972} having realized, after due consideration, that his initial declared preference in a decision problem presented by \citet{allais_so-called_1979} was mistaken.

It is well known that shallow preferences do not always match reflective ones.\commentYM{à quoi sert la phrase précédente ?}
Indeed, forming reflective preferences is recognised as a difficult problem in many decision problems, contrary, by definition, to shallow preferences. For example, \citet[p.\ 6]{colell_microeconomic_1995} write, when presenting the preference relation that is the basis for most of their famous book about Microeconomic theory: “Introspection quickly reveals how hard it is to evaluate alternatives that are far from the realm of common experience. It takes work and serious reflection to find out one’s own preferences.” 
In the words of \citet[p.\ 9]{bell_descriptive_1988}, “Our decision makers are not economic automatons; they make mistakes, have remorse, suffer anxieties, and cannot make up their minds”\commentYM{cette citation me semble mal venue ici, car elle ne pose pas que la question du contraste shallow/reflective, mais aussi la question de savoir s'il faudrait toujours faire un calcul économique. Par exemple avoir des remords ne me semble pas forcément une mauvaise chose} (cited by \citet[p.\ 87]{deparis_etude_2012}).
Indeed, it is to be expected that some behavior highlighted by psychologists \citep{kahneman_thinking_2012}, such as framing effects, would be rejected by the individuals themselves when pondering their decisions. \commentYM{formulation ambigue : on dirait que tu dis que les gens nieraient que les framing effects existent}

@@

\subsection{Existing approaches for capturing reflective preferences}
\subsubsection{Laundering preferences}
An approach much discussed recently is the one of laundering preferences, also called preference purification \citep{bernheim_beyond_2009, rubinstein_eliciting_2011, infante_preference_2016}: having collected shallow preference data (from the field or from interviews), laundering preferences consists in transforming them, using a formal process, in such a way that they satisfy axioms generally considered as required by rationality, or keeping only preference data collected in some situations considered appropriate to determine welfare-relevant preferences. For example, \citet{bleichrodt_making_2001} transform observed preferences in order to obtain a classical Expected Utility model; \citet{bernheim_beyond_2009} propose to exclude situations where the individual seems (by objective criteria) unable to process information considered (by the theorist) as relevant information. \commentYM{les deux mécanismes évoqués sont très différents, et ont des enjeux différents. Je pense qu'il serait utile de les distinguer, et je pense qu'il faudrait en dire plus sur la nature de la transformation et du filtrage, sans quoi la discussion des ces approches est trop superficielle}

>>>>

\section{Introduction} 
Decision theoretic approaches in economy divide into (at least) two sub-fields. Normative approaches aim at prescribing decision behavior; descriptive approaches aim at describing decision behavior. This partition is not always (depending on the authors) considered a complete cover of decision theoretic approaches in economy, but it is definitely considered disjoint.

As a result, normative approaches are difficult to validate: being considered as non-descriptive in their goal, normative claims cannot be submitted to empirical tests that would permit to verify or falsify them.
Instead, the validity of a normative approach is to be determined by examining the convincingness power of the normative axioms on which it rests.
This validity criterion is unsatisfactory, as it is unable to settle disputes between experts when they arise. In supplement, agreement between experts in decision theory is arguably not enough to ground the practical applicability of some normative axioms: it is desirable to check that decision makers to which the prescription is addressed agree with the particular instanciation of these axioms in the specific decision problem under study. But no systematic, formally defined procedure has been proposed to proceed with this check, to the best of my knowledge. 
In practical applications, an expert may informally discuss the axioms with the decision maker in order to check endorsement, either in general, or by discussing violations on a case by case basis when they are observed during an interactive elicitation session. 
The resulting state of conviction arguably depends in part on the rethorical ability of the expert, not solely on the normative strength of the axioms. It may lead to a decision maker accepting a decision procedure considered unsound by many experts in the field, such as AHP. 
This problem is compounded by the enormous constraining power gained when axioms interact, as the celebrated theorem of Arrow shows dramatically: axioms that appear innocuous when taken alone may precipitate undesirable results when combined.

By contrast, this difficulty of validation is much alleviated with descriptive approaches. Such approaches can be validated by comfronting their predictions to empirical data. When two descriptive theories disagree on their predictions, it is often possible to show that at least one of them is wrong with a suitable experiment. 
Alàs, descriptive approaches as currently existing in decision theory are not suitable to obtain valid prescriptions.

This article defines decision theories, called theories of \ac{DJ}, that are both normative and descriptive: they claim to be able to yield valid prescriptions, and to allow confrontation to empirical data.

It is by no means my intent to claim that descriptive approaches never face validation difficulties. Experimental tests to arbitrate between theories may not be feasible in practice, descriptions are not always solely about the future, falsificationist approaches in turn raise epistemological difficulties, and so on. Correspondingly, the possible confrontation to empirical data of the normative and descriptive theories defined here will likely not suffice to close every discussion about validation. My less naïve claim, that should suffice to motivate the present endeavor, is that it is often a good thing for a theory to make empirically testable claims, from the point of view of the possibilities of solving disputes and validating theories.

As the plural indicates, this article defines a kind of theories, called theories of \ac{DJ}, not specific theories of \ac{DJ}. Specific theories of \ac{DJ} are tailored to a set of situations and yield claims relevant to those situations. For example, a specific theory of \ac{DJ} may be tailored to situations of choice between lotteries involving money. This article adopts a methodological point of view. It proposes a definition of what theories of \ac{DJ} should look like, and proves claims that hold for any specific theory of \ac{DJ}, showing that such theories are indeed normative and descriptive.

\Cref{sec:approaches} defines more precisely what is meant here with normative and descriptive theories. These definitions permit to readily see why the apparent impossibility of a claim being both normative and descriptive is an illusion, thereby enabling a solution to what could be called the “normative against descriptive” gap.

\section{Species of theories}
\label{sec:approaches}
For this article, a \emph{normative decision theory} is a theory that, when applied in the situations it self-describes as applicable to, claims to yield valid prescriptions. I consider a prescription in the wide sense: recommending against some behavior, or recommending conditionally, also counts as prescribing.

A \emph{descriptive decision theory} is a theory that, when correctly applied in the situations it self-describes as applicable to, claims to predict the decision behavior of a given set of individuals. For this article, a claim counts as a prediction only when it is in principle possible to show that it is false using empirical data, thus predictions only apply to empirically observable data.

I will simply write normative theory and descriptive theory from now on to mean normative and descriptive decision theories.

For example, the \ac{SEU} theory of Savage is a normative theory. It may be used to recommend conditionally some behavior: “if you choose this lottery rather than that lottery, then you ought to \emph{not} choose this other lottery rather than that other lottery”.
Examples of descriptive theories include the many variants of Prospect theory.

These definitions avoid to define the concept of theory itself, for defining precisely such a fundamental concept raises serious difficulties. Is Prospect theory in general a theory, or only some of its variants, sufficiently precisely defined? Can a theory be extremely specific, or is some level of generality required to deserve being called a theory? Must a theory consist in a non-trivial assembly of laws, some being deducible from others? It is not necessary for the purpose of the present discussion to take position on these issues, rather, the focus is on the distinction between normative and descriptive approaches. An intuitive understanding of the concept of theory will suffice.
Similarly, I sidestep insuperable difficulties pertaining to capturing in a short definition the concept of a descriptive theory in general: some descriptive theories may propose to describe without predicting, for example (historical decisions or trends come to mind); such endeavor simply does not enter the scope of this discussion, and I make no claim about its legitimacy to be called a descriptive theory in general.

I say that a normative theory is also a descriptive theory when the prescriptions of the normative theory, in supplement of claiming to be valid prescriptions (in the situations that the theory claim to be applicable to), also claim to predict the decision behavior of a given set of individuals.

Another contrast can now be introduced: the contrast between shallow and reflexive behavior. \emph{Shallow behavior} is defined as the everyday, usual behavior of an individual. 
This is the behavior whose observation defines what \citet[p.\ 16]{von_neumann_theory_2004} describe as the “intuitive” preferences of an individual, those originating from an “immediate sensation” (cited in \citet{fishburn_retrospective_1989}).
\emph{Reflexive behavior}, by contrast, is defined as the behavior that enacts informed, pondered decisions, those that would be taken after due consideration.
% an individual has consciously adopted after having pondered everything that is possibly relevant to the decision. 
%This pair of definitions is, again, vague and unsatisfactory for a deep philosophical discussion, but hopefully sufficient for the present one. 
%What is meant by “pondering about everything that is possibly relevant to the decision” will be defined more precisely later; an intuitive understanding of this phrase will do at this stage.
This distinction draws on our capacity, as human beings, to think about our shallow preferences and consider some of them as mistaken \citep{frankfurt_freedom_1971}. 
A famous example of such a situation is the one of \citet[pp.\ 101--103]{savage_foundations_1972} having realized, after due consideration, that his initial declared preference in a decision problem presented by \citet{allais_so-called_1979} was mistaken.

This contrast raises a new distinction, which appears by simple combination of the just given definitions: a \emph{shallow descriptive theory} is a descriptive theory that claims to predict the shallow behavior of individuals; a \emph{reflexive descriptive theory} is one that claims to predict the reflexive behavior of individuals.

The descriptive theories proposed in decision theory, in the sense defined here, are exclusively shallow descriptive theories, to the best of my knowledge. (\Cref{sec:laundering} discusses the theories that launder preferences.)
%This need not be so. Indeed, this article introduces descriptive theories about reflexive behavior.
This observation permits to understand what could be called the “normative versus descriptive gap”, implicit in the decision theoretic literature: because normative theories aim to prescribe, it is asked of them that they do not yield recommendations that precisely, systematically match shallow behavior. Given the numerous results in experimental psychology showing that shallow behavior is undefensible in many situations, a normative theory whose prescriptions would systematically match shallow behavior would be suspicious, to say the least. It is an immediate consequence of this minimal requirement that a normative theory cannot also be a shallow descriptive theory. 
As \citet{hausman_preference_1994} illustrate with an example of a person drinking a fatal poison while erroneously thinking that it is water, shallow behavior is not sufficient to inform normative economics.
The “normative versus descriptive gap” comes from an implicit identification of descriptive theories with shallow descriptive theories.

This gap can thus be overcome by considering reflexive descriptive theories, which is the path this article adopts.
It defines theories that are both normative and descriptive: they attempt to deliver valid prescriptions, and these prescriptions are claimed to predict reflexive behavior.

%This is done by defining on the one hand a non-observable set, containing those propositions that are valid prescriptions; and on the other hand an observable set, containing those propositions decisively defended by arguments, as judged by a given individual. This latter set is called the individual’s \ac{DJ}.
%A normative axiom, that any theory of \ac{DJ} rests on, links these two sets by stating that what is in someone’s \ac{DJ} is a valid prescription.
%This justifies the normative aspect of the theories. 
%The theory yields descriptive claims thanks to the

\section{Previous Motivation}
It is well known that shallow preferences do not always match reflective ones.\commentYM{à quoi sert la phrase précédente ?}
Indeed, forming reflective preferences is recognised as a difficult problem in many decision problems, contrary, by definition, to shallow preferences. For example, \citet[p.\ 6]{colell_microeconomic_1995} write, when presenting the preference relation that is the basis for most of their famous book about Microeconomic theory: “Introspection quickly reveals how hard it is to evaluate alternatives that are far from the realm of common experience. It takes work and serious reflection to find out one’s own preferences.” 
In the words of \citet[p.\ 9]{bell_descriptive_1988}, “Our decision makers are not economic automatons; they make mistakes, have remorse, suffer anxieties, and cannot make up their minds”\commentYM{cette citation me semble mal venue ici, car elle ne pose pas que la question du contraste shallow/reflective, mais aussi la question de savoir s'il faudrait toujours faire un calcul économique. Par exemple avoir des remords ne me semble pas forcément une mauvaise chose} (cited by \citet[p.\ 87]{deparis_etude_2012}).
Indeed, it is to be expected that some behavior highlighted by psychologists \citep{kahneman_thinking_2012}, such as framing effects, would be rejected by the individuals themselves when pondering their decisions. \commentYM{formulation ambigue : on dirait que tu dis que les gens nieraient que les framing effects existent}

@@

\subsection{Existing approaches for capturing reflective preferences}
%Yet, classical methods to obtain preference models in economics are either formally defined on the basis of observable data that do not describe reflective preferences, or are obtained by other means than observable preferential data and systematic reproducible formal procedures. \commentYM{je devine ce que tu veux dire, mais je ne comprends pas la dernière phrase. Il me semble important que tu t'expliques ici, car je pense que le lecteur ici ne peut pas comprendre la niche dans laquelle s'inscrit ton projet}

\subsubsection{Laundering preferences}
\label{sec:laundering}
An approach much discussed recently is the one of laundering preferences, also called preference purification \citep{bernheim_beyond_2009, rubinstein_eliciting_2011, infante_preference_2016}: having collected shallow preference data (from the field or from interviews), laundering preferences consists in transforming them, using a formal process, in such a way that they satisfy axioms generally considered as required by rationality, or keeping only preference data collected in some situations considered appropriate to determine welfare-relevant preferences. For example, \citet{bleichrodt_making_2001} transform observed preferences in order to obtain a classical Expected Utility model; \citet{bernheim_beyond_2009} propose to exclude situations where the individual seems (by objective criteria) unable to process information considered (by the theorist) as relevant information. \commentYM{les deux mécanismes évoqués sont très différents, et ont des enjeux différents. Je pense qu'il serait utile de les distinguer, et je pense qu'il faudrait en dire plus sur la nature de la transformation et du filtrage, sans quoi la discussion des ces approches est trop superficielle}
